---
title: "AI/ML SOS IDK"
subtitle: "A Beginner-Friendly Machine Learning Workflow in R"
format:
  html:
    toc: true
    toc-title: "On deck"
    toc-location: right
    code-fold: false
    code-copy: true
    code-tools: false
    embed-resources: true
    theme: cosmo
    css: styles.css
    fig-align: center
    fig-width: 8
    fig-height: 5
---

## Can we predict the NWSL playoffs? &#9917;

Before an NWSL season is over, can we predict which teams will make the playoffs?

We're going to answer that question with a random forest.

But the model isn't really the point.

**The workflow is.**

> The algorithm can be interesting. The workflow should be boring.

```{r}
#| label: setup
#| message: false
#| warning: false

library(tidymodels)
library(readr)
library(dplyr)
library(ggplot2)
library(knitr)

set.seed(20260914) 
```

## 1. Load the data

We've already separated our data chronologically:

- **Training:** 2016–2024
- **Testing:** 2025
- **Current season:** 2026

```{r}
#| label: load-data

train <- read_csv("data/training_data.csv", show_col_types = FALSE)
test <- read_csv("data/test_data.csv", show_col_types = FALSE)
current <- read_csv("data/current_2026_60.csv", show_col_types = FALSE)
```


## 2. Take a look

A quick sanity check before we build anything: how big are the datasets, and how balanced is the outcome we're trying to predict?

```{r}
#| label: inspect-data

tibble(
  Dataset = c("Training (2016–2024)", "Testing (2025)", "Current (2026)"),
  Rows = c(nrow(train), nrow(test), nrow(current)),
  `Team-seasons` = c(
    n_distinct(train$team_season_id),
    n_distinct(test$team_season_id),
    n_distinct(current$team_season_id)
  )
) |>
  kable(align = "lrr")
```

Each row is a snapshot of one team at one point in a season. Snapshots are taken at 20%, 40%, 60%, and 80% of the way through the schedule, so a single team-season shows up multiple times in the data.

```{r}
#| label: inspect-outcome

train |>
  count(made_playoffs, name = "Snapshots") |>
  kable(
    align = "lr",
    caption = "Training label balance"
  )
```

Roughly balanced — no rare-class problem to worry about.

Here is a small snapshot of the modeling data. Each row is one team's state at a different point in the season; the columns show a few predictors and the outcome we want to predict.

<!-- Presentation-only table; remove this chunk before creating the code handout. -->

```{r}
#| label: data-snapshot

train |>
  filter(
    season == 2016,
    team %in% c("Portland Thorns FC", "Seattle Reign FC")
  ) |>
  arrange(team, checkpoint_pct) |>
  transmute(
    Season = season,
    Team = team,
    `Season completed` = scales::percent(checkpoint_pct, accuracy = 1),
    `Points per game` = round(points_per_game, 2),
    Wins = wins,
    `Made playoffs` = made_playoffs
  ) |>
  kable(
    align = "l",
    caption = "Example rows from the training data",
    table.attr = 'class="data-snapshot"'
  )
```

## 3. Preprocess with a recipe

First, we'll make sure our outcome is treated as a categorical variable.

```{r}
train <- train |>
  mutate(made_playoffs = factor(made_playoffs, levels = c("Yes", "No")))

test <- test |>
  mutate(made_playoffs = factor(made_playoffs, levels = c("Yes", "No")))
```

Next we'll define the predictors and transformations using the **recipes** package. Some features are created inside the recipe so the same transformations are applied consistently to future data.

<!-- A recipe is a reproducible set of instructions for preparing data for a model. Why not dplyr? Then you have to remember to perform exactly the same transformations on the test data and the 2026 data. With a recipe, those same steps are automatically applied to new data later. -->

```{r}
playoff_recipe <- recipe(
  made_playoffs ~
    points +
    wins +
    goal_diff +
    games_played +
    last_5_points +
    last_5_goal_diff +
    points_rank_pct +
    points_gap_to_cutline +
    playoff_share +
    season_pct,
  data = train
) |>
  step_mutate(
    points_per_game = points / games_played,
    win_pct = wins / games_played,
    goal_diff_per_game = goal_diff / games_played
  ) |>
  step_rm(points, wins, goal_diff, games_played) |>
  step_zv(all_predictors())
```


## 4. Specify the model

Now we'll tell tidymodels what kind of model we want to fit. This is the **parsnip** part of tidymodels — a single, consistent interface for describing models regardless of which engine ends up running them.

```{r}
rf_spec <- rand_forest(
  trees = 500
) |>
  set_engine("ranger") |>
  set_mode("classification")
```

<!-- We're fitting a random forest with 500 trees, using ranger as the engine, and classification because our outcome is yes or no. The model specification describes what we want to fit, but it does not fit anything yet. -->


## 5. Build the workflow

A **workflows** object bundles the preprocessing steps and model specification together.

```{r}
playoff_workflow <- workflow() |>
  add_recipe(playoff_recipe) |>
  add_model(rf_spec)
```

<!-- This is the key tidymodels idea: preprocessing and modeling travel together. -->


## 6. Fit the model

Now we can fit the entire workflow to our historical training data.

```{r}
playoff_fit <- playoff_workflow |>
  fit(data = train)
```


## 7. Evaluate on 2025

Now we'll test the fitted workflow on the held-out 2025 season.

We'll make our prediction 60% of the way through the season — early enough that the playoff field isn't completely settled yet.

```{r}
#| label: evaluate-2025

test_60 <- test |>
  filter(checkpoint_pct == 0.6)

test_predictions <- predict(
  playoff_fit,
  new_data = test_60,
  type = "prob"
) |>
  bind_cols(
    predict(playoff_fit, new_data = test_60)
  ) |>
  bind_cols(
    test_60 |>
      select(team, made_playoffs)
  ) |>
  arrange(desc(.pred_Yes))
```

The **RF playoff score** shows how strongly the random forest leans toward a playoff outcome. The class prediction just applies a 50% cutoff to that score.

```{r}
#| label: evaluate-2025-table

test_predictions |>
  transmute(
    Team = team,
    Actual = made_playoffs,
    Predicted = .pred_class,
    `RF playoff score` = scales::percent(.pred_Yes, accuracy = 1)
  ) |>
  kable(align = "lccr")
```

```{r}
#| label: evaluate-2025-metrics

accuracy(
  test_predictions,
  truth = made_playoffs,
  estimate = .pred_class
) |>
  transmute(Metric = .metric, Value = scales::percent(.estimate, accuracy = 1)) |>
  kable(align = "lr")

conf_mat(
  test_predictions,
  truth = made_playoffs,
  estimate = .pred_class
)
```

Yes, that says **100%**. No, we have not solved soccer. This is one held-out season at one checkpoint — 14 teams, all classified correctly. Give the model a whole different league, a different year, or an earlier checkpoint and it will start missing. What we can say is that at the 60% mark of the 2025 NWSL season, the workflow got every call right.


## 8. Refit with all historical data

We held out 2025 so we could evaluate our model on data it had never seen.

Now that we've done that, we'll refit the same workflow using all of our completed historical seasons.

```{r}
#| label: refit-final

all_history <- bind_rows(train, test)

final_fit <- playoff_workflow |>
  fit(data = all_history)
```

<!-- Nothing about our workflow changed. We're simply giving it more historical data to learn from. -->


## 9. Predict the 2026 playoffs

Finally, we'll apply our fitted workflow to the 2026 season as it stands today.

```{r}
#| label: predict-2026

current_predictions <- predict(
  final_fit,
  new_data = current,
  type = "prob"
) |>
  bind_cols(
    predict(final_fit, new_data = current)
  ) |>
  bind_cols(
    current |>
      select(team)
  ) |>
  arrange(desc(.pred_Yes))
```

> **A note on the numbers:** the RF playoff score shows how strongly the random forest leans toward a playoff outcome. It's a model score, not calibrated real-world playoff odds.

```{r}
#| label: predict-2026-table

current_predictions |>
  transmute(
    Team = team,
    `RF playoff score` = scales::percent(.pred_Yes, accuracy = 1)
  ) |>
  kable(align = "lr")
```

Washington Spirit lands around **99%** — basically every tree in the forest is landing on "playoffs." The model is very strongly Team Playoffs for Washington. Orlando sits around **47%**, leaning no but much more of a coin-flip.

```{r}
#| label: fig-playoff-scores
#| fig-cap: "Random forest playoff scores for the 2026 NWSL season. Model scores, not calibrated playoff odds."
#| fig-width: 8
#| fig-height: 6

spirit_green <- "#003C34"  # deep teal
lemon_venom  <- "#E8F82C"  # vibrant yellow, Washington highlight
spirit_black <- "#0A0A0A"
muted_bar    <- "#B8BEC2"  # quiet gray for the rest of the league

playoff_scores <- current_predictions |>
  transmute(
    team,
    playoff_score = .pred_Yes,
    is_spirit = team == "Washington Spirit"
  ) |>
  arrange(playoff_score) |>
  mutate(team = factor(team, levels = team))

ggplot(
  playoff_scores,
  aes(x = playoff_score, y = team, fill = is_spirit, color = is_spirit)
) +
  geom_col(width = 0.72, linewidth = 0.6) +
  geom_vline(
    xintercept = 0.5,
    linetype = "dashed",
    color = "#9CA3AF",
    linewidth = 0.5
  ) +
  geom_text(
    aes(label = scales::percent(playoff_score, accuracy = 1)),
    hjust = -0.15,
    size = 3.6,
    color = spirit_black
  ) +
  scale_x_continuous(
    labels = scales::percent_format(accuracy = 1),
    limits = c(0, 1.08),
    breaks = seq(0, 1, 0.25),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_fill_manual(
    values = c(`TRUE` = lemon_venom, `FALSE` = muted_bar)
  ) +
  scale_color_manual(
    values = c(`TRUE` = spirit_green, `FALSE` = muted_bar)
  ) +
  labs(
    title = "So... who's making the playoffs?",
    subtitle = "Random forest playoff scores for the 2026 NWSL season",
    caption = "Model scores, not calibrated real-world playoff odds.",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", color = spirit_black, size = 16),
    plot.subtitle = element_text(color = spirit_green, margin = margin(b = 12)),
    plot.caption = element_text(
      color = "#6B7280",
      hjust = 0,
      margin = margin(t = 10),
      face = "italic"
    ),
    axis.text.y = element_text(color = spirit_black),
    axis.text.x = element_text(color = "#6B7280"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_line(color = "#EAEAEA"),
    plot.margin = margin(10, 20, 10, 10)
  )
```

<!-- We just built a random forest model and predicted the 2026 NWSL playoffs. But notice what most of our code was actually about: structuring the data, defining preprocessing, keeping the model and preprocessing together, evaluating on held-out data, and applying the workflow to new data. The random forest itself was three lines of code. That's the point. The algorithm can be interesting. The workflow should be boring. -->

